Tensor denoising and completion based on ordinal observations

### 2020-02-02 ICML/chanwoo/ICMLdraft.tex
Add files via upload

\subsection{Rank selection}
Algorithm \ref{alg} takes  the rank $\bm r$ as an input variable. In practice, the rank $\bm r$ is hardly known. Estimating an appropriate rank $\bm r$ from a given tensor is an important issue. We suggest to use Bayesian Information Criterion(BIC) to choose the rank.
\begin{equation}
    \label{bic}
    \begin{aligned}
        \hat{\bm r} &= \argmin_{\bm r\in \mathbb{N}_+^K} BIC(\bm r)\\
        &=\argmin_{\bm r\in \mathbb{N}_+^K}\bigg[-2\mathcal{L}_{\mathcal{Y},\Omega}(\hat\Theta(\bm r),\hat\mb(\mr)) +
        p_e(\mr)\log\big(\prod_{k\in[K]} d_k\big)\bigg],
    \end{aligned}
\end{equation}
where $\hat\Theta(\bm r),\hat\mb(\mr)$ is a maximum likelihood estimate given the rank $\bm r$, and $p_e(\rm)\eqdef \sum_k(d_k-r_k)r_k = \prod_k r_k$ is the effective number of free parameters in the model. We select the rank $\hat{\bm r}$ that minimizes BIC value through the grid search method.
\section{Real-world Data Applications}
In this section, we apply our ordinal tensor decomposition method to two real-world datasets of ordinal tensors. In the first application, we use our model to analyze an ordinal tensor consisting of structural connectivity patterns among 68 brain regions for 136 individuals from Human Connectome Project (HCP) \citep{geddes2016human}. In the second application, we perform tensor completion from the data with missing values. The data tensor records the ratings from scale 1 to 5 of 42 users to 139 songs on 26 contexts \citep{baltrunas2011incarmusic}.
\subsection{Human Connectome Project (HCP)}
The human connectome project (HCP) is a $68\times 68 \times 136$ tensor where the first two modes have 68 indices representing brain regions and the last mode has 136 indices meaning individuals. All the individual images were preprocessed following a standard pipeline \citep{zhang2018mapping}, and the brain was parcellated to 68 regions of interest following the Desikan atlas \citep{desikan2006automated}. The tensor entries consist of $\{1,2,3\}$, the strength of fiber connections between 68 brain regions for each of the 136 individuals. We apply our ordinal tensor decomposition method with a logistic link function to the HCP data. The BIC result suggests $\bm r = (23,23,8)$  with $\mathcal{L}_\mathcal{Y}(\hat{\Theta},\hat{\mb}) = -216645.8$.

We check the performance of our model and compare with other methods: the continuous tensor decomposition method, the 1 bit-tensor completion method with probit link function and the 1 bit-tensor completion method with logistic link function. We use 5 folded cross validation method for the comparison. Specifically, we randomly split the tensor entries into 5 similar sized pieces and alternately use each piece of entries as a test data using other $80\%$ entries as a training data. The entries in the test data are encoded as missing and then predicted based on each method from the training data. In our model,  predict $y_\omega$ as $\hat{y}_\omega^{\text{(mode)}} = \argmax_\ell(g_\ell(
\hat\theta_\omega,\hat\mb)).$ Table \ref{CV} shows RMSE, MAE and error(false prediction rate), averaged over 5 test set results. We can check that our ordinal tensor decomposition model outperforms the other methods in all criteria.

\begin{table}[t]
\label{CV}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lcccr}
\toprule
 Method & MSE & MAD & MCR \\
\midrule
\begin{tabular}[c]{@{}c@{}}Oridnal tensor\\ Decomposition\end{tabular}  & 0.1503 & 0.1502& 0.1502\\ \\
\begin{tabular}[c]{@{}c@{}}Conti tensor\\ Decomposition\end{tabular} & 0.1603& 0.1600& 0.1598\\ \\
\begin{tabular}[c]{@{}c@{}}1bit-completion\\ (probit)\end{tabular}  & 0.3658 & 0.3632 & 0.3619\\
\begin{tabular}[c]{@{}c@{}}1bit-completion\\ (logistic)\end{tabular}  &  0.4514 & 0.4215 & 0.4066
\\
\bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\caption{Results of comparisons among 4 methods on the HCP data predicting the test data. Four methods are the ordinal tensor decomposition algorithm, the continuous tensor decomposition algorithm and the 1 bit tensor completion method with probit link function and logistic link function. Each method is evaluated by RMSE, MAE, error(false prediction rate).}
\end{table}

\subsection{InCarMusic recommendation system}
InCarMusic is a mobile application that offers music recommendation to passengers of cars based on contexts (Baltrunas et al. 2011). To be specific, the music recommendations can be different according to the mood of the driver or the traffic condition in this system. Our goal is to perform the tensor completion to the $42\times 139\times 26$ ordinal tensor and thereby we can offer context-specific music recommendation to users. The tensor entries consist of ordinal observations on the scale 1 to 5, the ratings of 42 users to 139 songs on 26 contexts and are encoded as NA for missing values. The number of missing values is 148,904 and the number of available values is 2,884.
We suggest three estimators $\hat y_\omega$ based on $(\hat \theta_\omega,\hat \mb)$:
\begin{itemize}
\item (Mean) $\hat y^{(\text{Mean})}_\omega=\sum_\ell \ell g_\ell(\hat \theta_\omega,\hat \mb) $;
\item (Median) $\hat y^{(\text{Median})}_\omega=\min\{\ell\in[L]\colon f_\ell(\hat \theta_\omega,\hat \mb)\geq 0.5\}$;
\item (Mode) $\hat y^{(\text{Mode})}_\omega=\argmax_\ell g_\ell(\hat \theta_\omega,\hat \mb) $
\end{itemize}
Note that, under the ordinal tensor model~\eqref{mainmodel}, the estimator $\hat y^{(\text{Mean})}_\omega$ minimizes $\mathbb{E}_{\hat{\theta}_\omega,\hat{\mb}}(y_\omega-y)^2$, the estimator $\hat y^{(\text{Median})}_\omega$ minimizes $\mathbb{E}_{\hat{\theta}_\omega,\hat{\mb}}|y_\omega-y|$
and the estimator $\hat y^{(\text{Mode})}_\omega $ minimizes $\mathbb{E}_{\hat{\theta}_\omega,\hat{\mb}}\mathds{1}_{\{y_\omega=y\}}$.
In contrast, these three estimators degenerate to a single estimator under the continuous-valued tensor decomposition model. We assess the accuracy using three metrics: Mean squared error (MSE), Mean absolute deviation (MAD), and Misclassification rate (MCR).

Results are averaged from 5 fold cross validation. {\color{red} add s.e. to the above table.}

>>>>

\subsection{Rank selection}
Algorithm \ref{alg} takes  the rank $\bm r$ as an input variable. In practice, the rank $\bm r$ is hardly known. Estimating an appropriate rank $\bm r$ from a given tensor is an important issue. We suggest to use Bayesian Information Criterion(BIC) to choose the rank.
\begin{equation}
    \label{bic}
    \begin{aligned}
        \hat{\bm r} &= \argmin_{\bm r\in \mathbb{N}_+^K} BIC(\bm r)\\
        &=\argmin_{\bm r\in \mathbb{N}_+^K}\bigg[-2\mathcal{L}_{\mathcal{Y},\Omega}(\hat\Theta(\bm r),\hat\mb(\mr)) +
        p_e(\mr)\log\big(\prod_{k\in[K]} d_k\big)\bigg],
    \end{aligned}
\end{equation}
where $\hat\Theta(\bm r),\hat\mb(\mr)$ is a maximum likelihood estimate given the rank $\bm r$, and $p_e(\rm)\eqdef \sum_k(d_k-r_k)r_k = \prod_k r_k$ is the effective number of free parameters in the model. We select the rank $\hat{\bm r}$ that minimizes BIC value through the grid search method.
\section{Real-world Data Applications}
In this section, we apply our ordinal tensor decomposition method to two real-world datasets of ordinal tensors. In the first application, we use our model to analyze an ordinal tensor consisting of structural connectivity patterns among 68 brain regions for 136 individuals from Human Connectome Project (HCP) \citep{geddes2016human}. In the second application, we perform tensor completion from the data with missing values. The data tensor records the ratings from scale 1 to 5 of 42 users to 139 songs on 26 contexts \citep{baltrunas2011incarmusic}.
\subsection{Human Connectome Project (HCP)}
The human connectome project (HCP) is a $68\times 68 \times 136$ tensor where the first two modes have 68 indices representing brain regions and the last mode has 136 indices meaning individuals. All the individual images were preprocessed following a standard pipeline \citep{zhang2018mapping}, and the brain was parcellated to 68 regions of interest following the Desikan atlas \citep{desikan2006automated}. The tensor entries consist of $\{1,2,3\}$, the strength of fiber connections between 68 brain regions for each of the 136 individuals. We apply our ordinal tensor decomposition method with a logistic link function to the HCP data. The BIC result suggests $\bm r = (23,23,8)$  with $\mathcal{L}_\mathcal{Y}(\hat{\Theta},\hat{\mb}) = -216645.8$.
First, we estimate latent parameters $\Theta$ from the full dataset and cluster the brain nodes based on $K$-means method from the estimated latent parameters. We decide the number of clusters as eight from the elbow method. The cluster result shows that most of the brain nodes fall into two biggest groups which can be represented as right-side brain and left-side brain.
The brain nodes, in each group except the above two biggest clusters, have the similar names and belong to the same location(right or left). For example, the fourth cluster group has the nodes having names such as \textit{l.postcentral, l.posteriorcingulate, l.pericalcarine and l.temporalpole,} all of which are encoded as ``SupraM$\_$L'' in the dataset. The brain clustering based on the latent estimator successfully catches regional traits of the brain nodes without external knowledge. More detailed procedure and result are provided in the supplements.

Secondly, we check the performance of our model and compare with other methods: the continuous tensor decomposition method\cite{filipovic2015tucker}, the 1 bit-tensor completion method\cite{ghadermarzy2018learning}. In 1 bit-tensor completion method, we use the probit link instead of the logistic link because the probit link has the better performance. We use stratified 5-folded cross validation method to assess the performances. Specifically, we randomly split the tensor entries into five pieces with the same size making sure that every piece has the similar proportion of ordinal observations.
We alternately use each piece of entries as a test dataset and use the other $80\%$ entries as a training dataset. The entries of the test dataset are encoded as missing and then predicted based on each method from the training dataset. In our model, We suggest three estimators $\hat y_\omega$ based on $(\hat \theta_\omega,\hat \mb)$:
\begin{itemize}
\item (Mean) $\hat y^{(\text{Mean})}_\omega=\sum_\ell \ell g_\ell(\hat \theta_\omega,\hat \mb) $;
\item (Median) $\hat y^{(\text{Median})}_\omega=\min\{\ell\in[L]\colon f_\ell(\hat \theta_\omega,\hat \mb)\geq 0.5\}$;
\item (Mode) $\hat y^{(\text{Mode})}_\omega=\argmax_\ell g_\ell(\hat \theta_\omega,\hat \mb) $
\end{itemize}
Notice that, under the ordinal tensor model~\eqref{mainmodel}, the estimator $\hat y^{(\text{Mean})}_\omega$ minimizes $\mathbb{E}_{\hat{\theta}_\omega,\hat{\mb}}(y_\omega-y)^2$, the estimator $\hat y^{(\text{Median})}_\omega$ minimizes $\mathbb{E}_{\hat{\theta}_\omega,\hat{\mb}}|y_\omega-y|$
and the estimator $\hat y^{(\text{Mode})}_\omega $ minimizes $\mathbb{E}_{\hat{\theta}_\omega,\hat{\mb}}\mathds{1}_{\{y_\omega=y\}}$.
We assess the accuracy using two metrics: Mean absolute deviation (MAD), and Misclassification rate (MCR).

We repeat stratified cross validation 10 times, average results and calculate standard deviations in Table \ref{table:CV}.
\begin{table}[H]
    \label{table:CV}
\scriptsize
\center
\begin{sc}
\begin{tabular}{ll|c|c}
\hline
\multicolumn{2}{c|}{Method}                       & MAD & MCR\\
\hline
\multirow{2}{*}{Our method} & $\hat y^{(\text{Median})}_\omega$ & \begin{tabular}[c]{@{}c@{}} 0.1607 (0.00046)\end{tabular} & \begin{tabular}[c]{@{}c@{}}0.1606 (0.00047)
\end{tabular} \\
&$\hat y^{(\text{Mode})}_\omega$         & \begin{tabular}[c]{@{}c@{}}0.1607  (0.00046)\end{tabular} & \begin{tabular}[c]{@{}c@{}}0.1606 (0.00047)\end{tabular} \\ \cline{1-2}
\multicolumn{2}{l|}{\begin{tabular}[c]{@{}l@{}}Cotinuous method\\~\cite{filipovic2015tucker}\end{tabular}} & \begin{tabular}[c]{@{}c@{}}0.2530  (0.00015)\end{tabular} & \begin{tabular}[c]{@{}c@{}}0.1599 (0.00023)\end{tabular} \\ \cline{1-2}
\multicolumn{2}{l|}{\begin{tabular}[c]{@{}l@{}}1bit Tensor Completion\\~\cite{ghadermarzy2018learning}\end{tabular}} & \begin{tabular}[c]{@{}c@{}}0.3566 (0.00102)\end{tabular} & \begin{tabular}[c]{@{}c@{}}0.3563 (0.00102)\end{tabular} \\ \hline
\end{tabular}
\end{sc}
\caption{Comparison of tensor completion performance on HCP data: values in the parentheses are standard deviation from 10 repetitions.}
\end{table}

\subsection{InCarMusic recommendation system}
InCarMusic is a mobile application that offers music recommendation to passengers of cars based on contexts (Baltrunas et al. 2011). To be specific, the music recommendations can be different according to the mood of the driver or the traffic condition in this system. Those kinds of external events that have influence on ratings on the musics are regarded as the contexts. Our goal is to perform the tensor completion to the $42\times 139\times 26$ ordinal tensor and thereby we can offer context-specific music recommendation to users. The tensor entries consist of ordinal observations on the scale 1 to 5, the ratings of 42 users to 139 songs on 26 contexts and are encoded as NA for missing values. The number of missing values is 148,904 and the number of available values is 2,884.
We perform 5 fold cross validation to assess the performance our model on InCarMusic data. The estimation methods and the procedure are the same as in HCP data analysis.
Results are averaged from 5 fold cross validation. {\color{red} add s.e. to the above table.}
